---
title: "Introduction to DiD"

knit: ( 
  function(input_file, encoding) {
    out_dir <- 'docs';
    
    rmarkdown::render(
    input       = input_file,
    encoding    = encoding,
    output_file = file.path(dirname(input_file), out_dir, 'intro.html') ) })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = T,
                      eval = T,
                      include = T,
                      message = FALSE,
                      warning = FALSE,
                      results = "hide")
```

# DiD: The basic principles

```{r preperations, echo = FALSE}
library(ggplot2)

knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```

## The Canonical Case
The canonical case is illustrated by 2 units at 2 points in time, giving a 2x2 matrix. One unit is treated, the other is not. The following graph illustrates this situation:

```{r canonical, fig.cap="The canonical case", warning = FALSE}
canon <- data.frame(id = paste0("Unit ",
                                rep(1:2, each = 2)),
                    time = rep(1:2, 2),
                    y = c(1, 2, rep(0, 2)),
                    Treated = c("No", "Yes", "No", "No"))

ggplot(canon, aes(x = time, y = y, color = id)) +
  geom_line() +
  scale_color_manual(values = c("red", "blue")) +
  scale_fill_manual(values = c("white", "grey50")) +
  geom_label(aes(label = y, fill = Treated)) +
  scale_y_continuous(limits = c(-1, 3), breaks = c(0,1,2) ) +
  theme_minimal()


```

In the canonical case, we can calculate ATT with the following formula:

$ATT = (Y_{1}^{t_{2}} - Y_{1}^{t_{1}}) - ( Y_{2}^{t_{2}} - Y_{2}^{t_{1}}) = (2 - 1) - (0 - 0) = 1 - 0 = 1$

In this situation, and similar situations with longer time-series and where treatment occurs for some units at the same point in time, the TWFE and DiD-estimator works well.

We can complicate the canonical case in three ways, each of which offers problems for the standard DiD-estimator/TWFE: 1) Treatment occurs at different points in time, 2) treatment effect is heterogeneous over time, and 3) units can reverse from the treatment status.

<!-- Parallel trends assumption: E[Y2(0)\|D1] - E[Y1(0)\|D1] = E[Y2(0)] -->

## The many problems

### Staggered adoption

Staggered adoption refers to the situation where units receive the treatment at different points in time. Staggered adoption does not cause any problem for the DiD-estimator/TWFE if the treatment effect is homogeneous (see [**Goodman-Bacon**](problems.html)).

Figure \@ref(fig:staggered) illustrates a situation with 3 units, staggered adoption of treatment, and homogeneous treatment effects.

```{r staggered, fig.cap="Staggered adoption"}
staggered <- data.frame(id = paste0("Unit ",
                                    rep(1:3, each = 6)),
                        time = rep(1:6, 3),
                        y = c(rep(2, 2), rep(3, 4),
                              rep(1, 3), rep(2, 3),
                              rep(0, 6)),
                        Treated = c(rep("No", 2), rep("Yes", 4),
                                    rep("No", 3), rep("Yes", 3),
                                    rep("No", 6)
                        ) 
)

ggplot(staggered, aes(x = time, y = y, group_by = id, color = id)) +
  geom_line() +
  scale_color_manual(values = c("red", "darkgreen", "blue")) +
  scale_fill_manual(values = c("white", "grey50")) +
  geom_label(aes(label = y, fill = Treated)) +
  scale_y_continuous(limits = c(-1, 4), breaks = c( 0:4 ) ) +
  theme_minimal()

```


### Heterogeneous treatment effect

The combination of staggered adoption and heterogeneous treatment effect _can_ cause the standard DiD-estimator/TWFE to assign negative ``weights'' to treated units (see [**Goodman-Bacon**](problems.html) and [**Jakiela**](problems.html)). If this happens, already treated units start to act as non-treated comparisons for other treated units.

Figure \@ref(fig:hetero) illustrates a situation with 3 units, staggered adoption of treatment and heterogeneous treatment effects.

```{r hetero, fig.cap="Staggered adoption with heterogeneous treatment effect"}
hetero <- data.frame(id = paste0("Unit ",
                                           rep(1:3, each = 6)),
                               time = rep(1:6, 3),
                               y = c(rep(2, 2), rep(3, 4),
                                     rep(1, 3), rep(4, 3),
                                     rep(0, 6)),
                               Treated = c(rep("No", 2), rep("Yes", 4),
                                           rep("No", 3), rep("Yes", 3),
                                           rep("No", 6)
                               ) 
)

ggplot(hetero, aes(x = time, y = y, group_by = id, color = id)) +
  geom_line() +
  scale_color_manual(values = c("red", "darkgreen", "blue")) +
  scale_fill_manual(values = c("white", "grey50")) +
  geom_line() +
  geom_label(aes(label = y, fill = Treated)) +
  scale_y_continuous(limits = c(-1, 5), breaks = c( 0:4 ) ) +
  theme_minimal()

```


### Reversal of treatment status

The combination of staggered adoption and heterogeneous treatment effect _can_ cause the standard DiD-estimator/TWFE to assign negative weights (and thus not really weights) to some data points (see [**Goodman-Bacon**](problems.html)).

Figure \@ref(fig:hetero) illustrates a scenario with staggered adoption, heterogeneous treatment effect, and reversal.


```{r reverse, fig.cap="Staggered adoption with heterogeneous treatment effect and reversal of treatment status"}
reverse <- data.frame(id = paste0("Unit ",
                                                   rep(1:3, each = 6)),
                                       time = rep(1:6, 3),
                                       y = c(rep(2, 1), rep(3, 2), rep(2, 2), 4,
                                             rep(1, 3), rep(4, 2), 3,
                                             rep(0, 6)),
                                       Treated = c(rep("No", 1), rep("Yes", 2), rep("No", 2), "Yes",
                                                   rep("No", 3), rep("Yes", 2), "No",
                                                   rep("No", 6) )
)

ggplot(reverse, aes(x = time, y = y, group_by = id, color = id)) +
  scale_color_manual(values = c("red", "darkgreen", "blue")) +
  scale_fill_manual(values = c("white", "grey50")) +
  geom_line() +
  geom_label(aes(label = y, fill = Treated)) +
  scale_y_continuous(limits = c(-1, 5), breaks = c( 0:4 ) ) +
  theme_minimal()
  
```


### Non-instantaneous treatment effects
When treatment effects are not instantaneous, it is possible that units that receives the treatment late are compared to units that have received the treatment and where the full effect of the first treatment has not yet manifested. This will bias the ATT towards 0.

Figure \@ref(fig:noninstantaneous) illustrates this situation. To understand why the early treated unit becomes a part of the comparison group for the late treated, see [**Goodman-Bacon**](problems.html). Naturally, this situation could also be combined with reversal of treatment effect, where the effect of reversal is also heterogeneous across units and time.


```{r noninstantaneous, fig.cap="Spillover of treatment effect from early treated into the comparison of the late treated"}

slow <- data.frame(id = c(rep("Early treated", 6),
                          rep("Late treated", 6),
                          rep("Never treated", 6)),
                        time = rep(1:6, 3),
                        y = c(rep(2, 2), 3, 4, 5, 5,
                              rep(1, 3), 2, rep(3, 2),
                              rep(0, 6)),
                        Treated = c(rep("No", 2), rep("Yes", 4),
                                    rep("No", 3), rep("Yes", 3),
                                    rep("No", 6)
                        ) 
)

ggplot(slow, aes(x = time, y = y, group_by = id, color = id)) +
  geom_line() +
  scale_color_manual(values = c("red", "darkgreen", "blue")) +
  scale_fill_manual(values = c("white", "grey50")) +
  geom_label(aes(label = y, fill = Treated)) +
  scale_y_continuous(limits = c(-1, 5), breaks = c( 0:5 ) ) +
  theme_minimal()


```

## Summary: Problems for TWFE

The potential problems with the standard TWFE estimator can be summarized in 2 points:

1. Treated units can receive negative treatment weights. This occurs because already treated units can enter into the equation as control units.

2. Rely on parametric assumptions that might not be correct.

The contributions offer two general approaches to adapt the DiD-estimator to these problems: a) a divide-and-conquer approach that splits the many-on-many comparison into many 2-by-2 comparisons and filter out those we should not use (e.g., comparisons between already treated and newly treated units), and  b) correct parameterization of the regression problem (e.g., estimating ATT for each treatment cohort). See [**Solutions**](solutions.html) for more details.

